---
title: "Getting Started with eiCalibrate"
author: "Loren Collingwood"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with eiCalibrate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**eiCalibrate** provides tools for empirical Bayes calibration of Dirichlet
priors in RxC ecological inference. In Voting Rights Act (VRA) litigation and
related racially polarized voting (RPV) analyses, ecological inference (EI)
estimates can be sensitive to the choice of prior. This package lets you:

1. **Simulate** benchmark election data with known ground truth
2. **Fit** ecological inference models via a unified interface (RxC Bayesian,
   iterative 2x2, Goodman's regression)
3. **Calibrate** Dirichlet concentration parameters against a known benchmark
4. **Sweep** prior sensitivity without ground truth
5. **Compare** methods side by side
6. **Cross-validate** via leave-one-out CV
7. **Visualize** all of the above

The core idea: use a calibration election where ground truth is available (e.g.,
from exit polls or a racially uncontested contest) to learn good hyperparameter
values, then apply those calibrated priors to the target election of interest.

## Installation

```{r install, eval = FALSE}
# From a local source directory:
devtools::install("/path/to/eiCalibrate")

# Or from GitHub:
devtools::install_github("lorenc5/eiCalibrate")
```

## Quick Start: Simulated Data

### Step 1: Simulate an Election

`simulate_election()` generates precinct-level data with known group-conditional
voting rates. This is useful for testing and benchmarking.

```{r simulate}
library(eiCalibrate)

dat <- simulate_election(
  n_precincts = 40,
  true_support = matrix(
    c(
      0.80, 0.20,
      0.15, 0.85
    ),
    nrow = 2, byrow = TRUE,
    dimnames = list(c("white", "black"), c("cand_A", "cand_B"))
  ),
  seed = 42
)

head(dat)
```

The true support matrix is stored as an attribute:

```{r truth}
(truth <- attr(dat, "true_support"))
```

### Step 2: Fit a Single Model

`fit_ei()` is the unified interface for all three EI methods. It accepts a
consistent set of column-name arguments.

**Goodman's ecological regression** (fast, no MCMC):

```{r goodman}
fit_good <- fit_ei(
  dat,
  method = "goodman",
  race_cols = c("white", "black"),
  cand_cols = c("cand_A", "cand_B"),
  totals_col = "total"
)

fit_good$estimates
```

**Iterative 2x2 EI** (via eiCompare):

```{r iter, message = FALSE, warning = FALSE}
fit_it <- fit_ei(
  dat,
  method = "iter",
  race_cols = c("white", "black"),
  cand_cols = c("cand_A", "cand_B"),
  totals_col = "total"
)

fit_it$estimates
```

**RxC Bayesian EI** (via eiPack --- slower, uses MCMC):

```{r rxc, message = FALSE, warning = FALSE, results = "hide"}
fit_rxc <- fit_ei(
  dat,
  method     = "rxc",
  race_cols  = c("white", "black"),
  cand_cols  = c("cand_A", "cand_B"),
  totals_col = "total",
  lambda1    = 4,
  lambda2    = 2,
  sample     = 10000,
  burnin     = 5000,
  thin       = 5,
  verbose    = FALSE
)
```

```{r rxc-show}
fit_rxc$estimates$point
```

### Step 3: Score Against Truth

```{r score}
score_estimates(fit_good$estimates, truth)
score_estimates(fit_rxc$estimates$point, truth)
```

## Calibrating the Prior

The key contribution of this package is calibrating the Dirichlet concentration
parameters $\lambda_1$ and $\lambda_2$ in `eiPack::ei.MD.bayes()`. The
calibration procedure sweeps a grid of $(\lambda_1, \lambda_2)$ values and
selects the pair that minimizes RMSE (or another loss) against a known benchmark.

### 1D Calibration (tied $\lambda_1 = \lambda_2$)

When you supply only `lambda_grid`, both concentration parameters are tied:

```{r calib-1d, message = FALSE, warning = FALSE, results = "hide"}
cal_1d <- calibrate_rxc(
  calib_data  = dat,
  calib_truth = truth,
  lambda_grid = c(0.1, 0.5, 1, 2, 4),
  sample      = 10000,
  burnin      = 5000,
  thin        = 5,
  verbose     = FALSE
)
```

```{r calib-1d-show}
cal_1d$summary
cal_1d$best_lambda1
```

### 2D Joint Calibration (independent $\lambda_1$, $\lambda_2$)

For finer control, supply separate grids. This explores the full
$\lambda_1 \times \lambda_2$ surface:

```{r calib-2d, eval = FALSE}
cal_2d <- calibrate_rxc(
  calib_data   = dat,
  calib_truth  = truth,
  lambda1_grid = c(0.1, 0.5, 1, 2),
  lambda2_grid = c(0.25, 0.5, 1, 2),
  n_reps       = 3,
  sample       = 25000,
  burnin       = 5000,
  thin         = 10,
  verbose      = TRUE
)

cal_2d$best_lambda1
cal_2d$best_lambda2
cal_2d$rmse_matrix
```

The `n_reps` argument averages over multiple MCMC runs at each grid point to
smooth out MCMC noise.

### Visualizing Calibration Results

```{r plot-calib-1d, fig.cap = "1D calibration: RMSE vs. lambda"}
library(ggplot2)
plot_calibration(cal_1d)
```

For 2D results, `plot_calibration()` produces a heatmap:

```{r plot-calib-2d, eval = FALSE}
plot_calibration(cal_2d)
```

## Sensitivity Analysis

When ground truth is unavailable, `sensitivity_sweep()` shows how estimates
change across prior strengths. This is useful for assessing whether conclusions
are robust.

```{r sweep, message = FALSE, warning = FALSE, results = "hide"}
sw <- sensitivity_sweep(
  dat,
  lambda_grid = c(0.1, 0.5, 1, 4),
  race_cols   = c("white", "black"),
  cand_cols   = c("cand_A", "cand_B"),
  totals_col  = "total",
  sample      = 10000,
  burnin      = 5000,
  thin        = 5,
  verbose     = FALSE
)
```

```{r plot-sweep, fig.cap = "Black support for Candidate A across prior values"}
plot_sensitivity(sw, cell = c("black", "cand_A"))
```

## Comparing Methods

`compare_methods()` runs multiple EI approaches on the same dataset and returns
a single data frame for easy comparison:

```{r compare, message = FALSE, warning = FALSE, results = "hide"}
comp <- compare_methods(
  dat,
  calibrated_lambda1 = cal_1d$best_lambda1,
  calibrated_lambda2 = cal_1d$best_lambda2,
  truth = truth,
  race_cols = c("white", "black"),
  cand_cols = c("cand_A", "cand_B"),
  totals_col = "total",
  sample = 10000,
  burnin = 5000,
  thin = 5,
  verbose = FALSE
)
```

```{r compare-show}
comp
```

```{r plot-compare, fig.cap = "Method comparison with ground truth"}
plot_comparison(comp)
```

## Leave-One-Out Cross-Validation

`loo_cv()` provides a principled way to compare methods without ground truth.
It holds out one precinct at a time, fits on the remaining data, predicts the
held-out precinct's vote share, and evaluates prediction error.

Goodman's regression is fast enough for a full LOO:

```{r loocv-goodman}
cv_good <- loo_cv(
  dat,
  methods    = "goodman",
  race_cols  = c("white", "black"),
  cand_cols  = c("cand_A", "cand_B"),
  totals_col = "total",
  verbose    = FALSE
)

cv_good$summary
```

For MCMC-based methods, LOO-CV is computationally expensive (one fit per
precinct per method). A typical call:

```{r loocv-full, eval = FALSE}
cv_all <- loo_cv(
  dat,
  methods = c("goodman", "rxc_default", "rxc_calibrated"),
  race_cols = c("white", "black"),
  cand_cols = c("cand_A", "cand_B"),
  totals_col = "total",
  calibrated_lambda1 = cal_1d$best_lambda1,
  calibrated_lambda2 = cal_1d$best_lambda2,
  sample = 10000,
  burnin = 5000,
  thin = 5,
  verbose = TRUE
)

cv_all$summary
```

### Visualizing CV Results

```{r plot-loocv-summary, fig.cap = "LOO-CV summary: RMSE and MAE by method"}
plot_loocv(cv_good, type = "summary")
```

```{r plot-loocv-precinct, fig.cap = "LOO-CV: per-precinct prediction errors"}
plot_loocv(cv_good, type = "precinct")
```

## Recommended Workflow

A typical analysis proceeds in stages:

1. **Identify a calibration election** where group-conditional truth is known
   (e.g., an uncontested race, exit poll data, or a BISG-validated contest).

2. **Run `calibrate_rxc()`** on the calibration data to find optimal
   $(\lambda_1^*, \lambda_2^*)$. Start with 1D calibration; move to 2D if
   $\lambda_2$ sensitivity matters.

3. **Validate with `loo_cv()`** to confirm calibrated priors improve predictive
   accuracy over the default.

4. **Run `sensitivity_sweep()`** on the target election to check robustness of
   conclusions across a range of priors.

5. **Run `compare_methods()`** on the target election to show convergence (or
   divergence) across Goodman, iterative EI, and RxC with both default and
   calibrated priors.

6. **Visualize** with `plot_calibration()`, `plot_sensitivity()`,
   `plot_comparison()`, and `plot_loocv()` for reports and expert testimony.

## Methodological Caution

Calibrating the prior is **only defensible** when:

- The calibration ground truth is genuinely credible.
- The calibration election has comparable dynamics to the target election (similar
  racial composition distribution, similar political context).
- The calibration and target elections are **separate**. Calibrating on the same
  election you analyze is circular.

Always report the calibration procedure transparently in expert reports and
litigation filings.

## Session Info

```{r sessioninfo}
sessionInfo()
```
